---
title: "From an expanded TCR clone to a shareable cell selection"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{From an expanded TCR clone to a shareable cell selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# The result first

This walkthrough follows one measured T-cell population through the whole
Viewer: a gene-defined expanded clone in Linked views, its sequence context in
HLA & TCR Motifs, and an anonymous link that reopens the same 293 cells for a
collaborator.

The source is the bundled `HLA & TCR` data set: 12,000 real CD8 T cells from
four donors, with transcriptomes, paired alpha/beta TCR contigs, raw dextramer
binder calls, and independently published donor HLA genotypes. For the raw-data
provenance, build pipeline, and source-specific caveats, read
`vignette("hla_tcr_antigen_selected")` first.

The case is versioned as two small files:

```{r case-files}
system.file(
  "extdata/examples/demo_hla_tcr_main_case.linked-view.json",
  package = "CerebroNexus"
)
system.file(
  "extdata/examples/demo_hla_tcr_main_case.expectations.json",
  package = "CerebroNexus"
)
```

The first is a normal Linked views configuration that a user can import. The
second is a machine-readable record of the expected counts and sequence family.

# The biological question

The deliberately narrow question is:

> Can an expanded, gene-defined TCR clone be located in the transcriptomic
> workspace, resolved into its TRB CDR3 sequence family, and shared as the same
> cell population with another viewer?

The selected clone uses the application's default `cloneCall = "gene"`
contract. Its `CTgene` identity is:

```text
TRAV27.TRAJ42.TRAC_TRBV19.None.TRBJ2-7.TRBC2
```

It contains 293 cells: 142 from donor1 and 151 from donor2. All 293 have the
raw `Flu-MP_Influenza` binder label, and the restricting allele named by that
reagent is present in both donors' published genotypes.

That observation is useful context, not proof of peptide specificity. A raw
binder label records staining by a reagent. A raw binder call is reagent
staining, not validated peptide specificity.

# Open the fixed selection

Launch the bundled Viewer from a source checkout:

```bash
Rscript -e 'pkgload::load_all("."); shiny::runApp("inst", port = 3838, launch.browser = TRUE)'
```

Then:

1. Select the **HLA & TCR** data set.
2. Open **Linked views**.
3. Press **Share selection**.
4. Under *Import or export a view*, press **Open JSON**.
5. Choose
   `inst/extdata/examples/demo_hla_tcr_main_case.linked-view.json`.

The imported configuration selects exactly 293 cells, colours them by `sample`,
opens the UMAP projection, and keeps the clonal layout in **Rank stack** mode.
The selected clone is rank six by the application's gene-level clone definition.

The portable JSON contains selected cell barcodes and view settings. It does not
contain the expression matrix, projection coordinates, image pixels, or local
paths; those remain in the matching `.crb` data set.

# Read the clone at the right level

The clonal panel and the motif network answer related but different questions:

| Layer | Identity | What this case contains |
|---|---|---|
| Linked views clone | `CTgene`, the alpha/beta V/J/C gene combination | 293 cells |
| Dominant sequence | TRB CDR3 amino-acid sequence | `CASSIRSSYEQYF` in 227 cells |
| Sequence family | equal-length TRB CDR3 nodes joined at Hamming distance 1 | 34 nodes, 627 cells |

The 293-cell CTgene clone contains ten distinct TRB CDR3 sequences. That is not
an inconsistency: `CTgene` intentionally groups by receptor genes and is coarser
than an exact amino-acid or nucleotide clonotype. Its dominant TRB sequence is
`CASSIRSSYEQYF`, but the clone must not be renamed as if all 293 cells carried
that exact sequence.

In Linked views, compare the highlighted cells in the clone rank stack with
their positions in UMAP. The coordinated highlight is descriptive: it shows
where this selected receptor-defined population lies in the expression
embedding; it does not by itself establish a transcriptional state or a
differential-expression result.

# Resolve the sequence family

Open **HLA & TCR Motifs**, use the TRB chain and the all-cell network, and locate
the anchor sequence `CASSIRSSYEQYF` in **Network data** or the feature picker.
The anchored component has:

- consensus `CASSxxxxxEQxF`;
- 34 distinct TRB CDR3 nodes;
- 627 cells in total;
- 308 cells from donor1, 318 from donor2, and 1 from donor4.

All ten TRB CDR3 sequences carried by the 293-cell CTgene clone belong to this
same component. The result is a measured pattern of sequence convergence within
this antigen-selected repertoire. The numeric motif label displayed by a given
graph build is intentionally absent from the case contract: motif numbers are
arbitrary, whereas the anchor and member sequences are stable identities.

# What the case supports

The evidence supports three statements:

1. the CTgene-defined clone is expanded across donor1 and donor2;
2. its ten TRB CDR3 sequences lie in one anchored Hamming-1 family;
3. that family is observed across donors in this selected cohort.

It does not establish any of the following:

- `CTgene` as an exact sequence-defined clonotype;
- peptide specificity from the raw dextramer binder label;
- HLA restriction from motif co-occurrence;
- a population-level HLA association from four antigen-selected donors;
- an unbiased estimate of repertoire frequency.

# Share the same cells

Return to **Linked views** with the imported selection still active:

1. press **Share selection**;
2. press **Create share link**;
3. copy the resulting URL;
4. open it in a private window or a browser with no Viewer login session.

The receiving Viewer should load the matching `HLA & TCR` population and restore
the same 293 selected cells and view settings. Creating and opening a selection
link does not require a login; authentication is reserved for administering
existing links.

A live share record expires after 90 days. The durable scientific object is
therefore the checked-in `demo_hla_tcr_main_case.linked-view.json`, not a URL in
documentation. After a link expires, import that JSON and create a fresh link.

# Rebuild and verify the case

The two JSON files are derived from the shipped `.crb`, not edited by hand:

```bash
Rscript data-raw/build_hla_tcr_main_case.R
```

The generator stops if the target CTgene clone, donor counts, dominant CDR3,
motif membership, or configuration fingerprint drifts. The corresponding test
independently reconstructs the selection and motif graph:

```bash
Rscript -e 'pkgload::load_all(".", quiet = TRUE); testthat::test_file("tests/testthat/test-hla-tcr-main-case.R", reporter = "summary")'
```

That pair provides both sides of a reproducible case: a human walkthrough and a
machine-checked biological contract.
